%=== CHAPTER ONE (1) ===
%=== INTRODUCTION ===

\chapter{Introduction}
\label{ch:introduction}

\begin{spacing}{2.0}
%\setlength{\parskip}{0.2in}

\section{Background and Context}
The rapid and ongoing expansion of global urban infrastructure has led to an unprecedented rise in vehicular traffic. As a result, this increased density has heightened the frequency of road traffic collisions and systemic congestion, thereby posing substantial risks to public safety and economic efficiency. In response to these escalating challenges, contemporary municipal authorities are progressively adopting Intelligent Transport Systems (ITS) to protect public infrastructure and optimise traffic management. Traditionally, traffic monitoring has depended heavily on intrusive hardware, such as inductive loop detectors, or manual human observation. However, these conventional methods present notable operational limitations; physical sensors are unable to furnish comprehensive spatial context, while human observation remains inherently vulnerable to fatigue-induced errors and delayed response times. To address these shortcomings, the integration of deep learning and computer vision technologies into existing Closed-Circuit Television (CCTV) networks has emerged as a highly effective, non-intrusive approach for collecting high-dimensional data concerning vehicle trajectories, interactions, and anomalies.

\section{Problem Statement}
Notwithstanding these advancements, the task of robust traffic anomaly detection continues to constitute a highly intricate challenge within the field of computer vision. Current object detection architectures, although highly effective at isolating individual vehicles under ideal conditions, often prove inadequate when faced with the chaotic and unstructured spatial dynamics characteristic of a physical collision. During such incidents, the visual profiles of vehicles tend to intersect, deform, or fragment, resulting in the failure of conventional bounding-box-based spatial trackers to maintain consistent identification.

Conversely, while complex spatiotemporal networks such as Transformer-based architectures, are capable of effectively modelling these interactions over time, they necessitate significant computational resources. This substantial processing requirement renders them largely impractical for real-time deployment on resource-constrained edge devices at the municipal level. Additionally, real-world environmental factors, including high-frequency camera oscillations caused by wind shear, adverse weather conditions, and varying illumination levels, markedly impair the reliability of isolated detection models. Systems that depend solely on either spatial tracking or deep learning classification frequently exhibit unacceptable false-positive rates or suffer from debilitating processing delays, thereby undermining their suitability for life-saving early-warning applications.

\section{The Proposed Hybrid Architecture}
To address this technological disparity, the present research introduces an innovative, hybrid vision-based framework devised explicitly for real-time traffic anomaly detection. The core advancement of this architecture resides in the rigorous separation of spatial vehicle tracking from kinetic anomaly classification. Rather than utilising a single, monolithic neural network to interpret both positional data and severity of impact, the system adopts an asynchronous, dual-threaded producer-consumer pipeline engineered to optimise hardware concurrency.

The primary spatial thread employs a pre-trained YOLOv1 model in conjunction with the BoT-SORT algorithm to establish and sustain persistent vehicle identities. By utilising Inverse Perspective Mapping (IPM) and Kalman filtering, the system projects two-dimensional pixel coordinates into a three-dimensional metric space, thereby enabling highly stable, real-time speed estimation. Simultaneously, a secondary kinetic thread assesses the unstructured physical motion within the scene employing the Farnebäck algorithm for dense optical flow. This raw kinetic data is encoded into a spatiotemporal Hue-Saturation-Value (HSV) tensor and subsequently classified using a custom-trained ResNet-18 Convolutional Neural Network (CNN).

Beyond the core detection capabilities, the pipeline is enhanced with programmatic solutions intended to address practical deployment constraints. These encompass mathematical camera stabilisation employing the Oriented FAST and Rotated BRIEF (ORB) algorithm to mitigate environmental camera shake, as well as automated Gaussian obfuscation to ensure strict adherence to the General Data Protection Regulation (GDPR) by disabling Personally Identifiable Information (PII) at the source.

\section{Aims and Objectives}
The primary objective of this research is to develop a highly accurate and computationally efficient prototype capable of sustaining a throughput of sixty frames per second on standard consumer-grade hardware. The system aims to integrate the deterministic logic associated with rule-based spatial tracking with the probabilistic pattern recognition offered by deep learning.

To this end, the network was meticulously trained and evaluated using a mathematically balanced synthesis of the Car Accident Detection and Prediction (CADP) and Traffic Anomaly Detection (TAD) benchmark datasets. In addressing the inherent class imbalance characteristic of traffic data where normal driving vastly exceeds incidents such as collisions, a Focal Loss function was employed to compel the network to dynamically prioritise complex kinetic boundary scenarios over more readily classified traffic patterns.

Ultimately, the goal is to achieve near-instantaneous detection latency for Time-to-Accident (TTA), thereby establishing a reliable foundation for automated municipal alert systems.

\section{Research Questions}
To assess the effectiveness and practical feasibility of the proposed hybrid architecture, this study systematically investigates the following primary research questions:
\begin{itemize}
    \item \textbf{RQ1:} Is it possible to effectively detect traffic abnormalities, such as collisions or abrupt stops, using a hybrid vision-based architecture that integrates YOLO vehicle monitoring with 3D metric speed estimation and dense optical flow?
    \item \textbf{RQ2:} What are the primary computational and environmental challenges of applying computer vision to kinematic speed-based anomaly identification across varied lighting and weather conditions?
\end{itemize}

\section{Structure of the Thesis}
The remainder of this thesis is organised as follows. Chapter 2 offers a comprehensive Literature Review, examining the development of object detection, multi-object tracking, speed estimation, and spatiotemporal optical flow analysis within the context of Intelligent Transportation Systems (ITS). Chapter 3 details the Methodology, describing the software architecture, data engineering pipelines, mathematical transformations, and the dual-threaded integration of the hybrid system. Chapter 4 presents the Analysis, Results, and Discussion, wherein the framework is assessed using quantitative classification metrics as well as qualitative real-world constraints. Finally, Chapter 5 concludes the study by summarising the key contributions to knowledge, acknowledging the inherent limitations of the system, and providing strategic recommendations for future development.

\end{spacing}
%=== END OF CHAPTER ONE ===
\newpage


