%=== CHAPTER FIVE (5) ===
%=== Discussion ===

\chapter{Conclusions and Recommendations}
\label{ch:conclusion}
\begin{spacing}{2.0}

\section{Summary of the Research}
The exponential expansion of urban infrastructure necessitates the development of resilient Intelligent Transport Systems (ITS) capable of mitigating the risks associated with traffic congestion and road accidents. Traditional computer vision methodologies have historically encountered challenges in balancing the requirement for highly accurate anomaly detection with the stringent latency constraints inherent to real-time edge computing, often relying on fragile object tracking techniques. To address this paradigm, this research has proposed, engineered, and evaluated a novel, solely spatiotemporal, vision-based framework. By structurally abandoning discrete spatial vehicle tracking, the system effectively assesses unstructured kinetic energy through Farnebäck dense optical flow, mapping this physical data into HSV tensors for pattern recognition by a custom-trained ResNet-18 classifier. This streamlined, deterministic approach circumvents the processing bottlenecks associated with sequential tracking, resulting in a comprehensive, real-world-ready prototype suitable for municipal traffic surveillance.
\section{Summary of Key Findings}
The empirical evaluation of the proposed architecture yielded highly competitive results, effectively addressing the primary research questions outlined at the commencement of this study.

In response to the initial research enquiry concerning the feasibility and efficacy of a purely spatiotemporal architecture, the system demonstrated exceptional reliability. Evaluated using a balanced dataset comprising 8,353 validation frames from the CADP and TAD benchmarks, the ResNet-18 classifier achieved an outstanding overall accuracy of 96.61\%. The utilisation of a Binary Cross-Entropy (BCE) loss function effectively delineated the boundary between normal and anomalous kinematics, resulting in an identical F1-score of 0.966 for both standard traffic conditions and severe accident scenarios. Furthermore, the implementation of an asynchronous, dual-threaded producer-consumer architecture entirely circumvented conventional processing bottlenecks, sustaining a real-time throughput of 60 frames per second on standard consumer-grade hardware. Attaining a Time-to-Accident (TTA) verification latency of merely 220 milliseconds further substantiates the system’s potential as a highly responsive early-warning mechanism.

In addressing the second research question concerning the computational and environmental challenges associated with vision-based detection, the evaluation emphasised the considerable benefits of ceasing the use of bounding-box mechanics. Dependence on spatial tracking is intrinsically vulnerable in scenarios characterised by heavy precipitation or casting shadows from bridges. Nonetheless, by solely mapping dense kinetic energy, the raw pixel displacement resulting from collisions was reliably identified irrespective of visual occlusion. This illustrates that assessing unstructured spatiotemporal energy confers substantially greater robustness to environmental degradation in comparison to conventional object detection frameworks.


\section{Contributions to Knowledge}
This study provides several significant contributions to the domain of automated traffic surveillance. Firstly, it validates a computationally efficient alternative to parameter-intensive Transformer architectures. By translating dense optical flow into HSV tensors and processing them through a modified ResNet-18 convolutional neural network, the framework achieves the complex spatiotemporal awareness traditionally associated with substantially larger models, whilst functioning within a computational capacity suitable for deployment on edge devices.

Secondly, the architectural choice to bypass spatial object detection sets an inherent privacy-by-design standard that is often overlooked within academic ITS research. As the framework focuses on isolating and assessing raw kinetic momentum rather than extracting high-resolution identifiers, such as licence plates or facial features, it thereby inherently aligns with strict data privacy regulations (such as GDPR) without the need for computationally intensive Gaussian obfuscation overlays.

\section{Research Limitations}
Despite the robust quantitative metrics attained, the current iteration of the framework possesses inherent limitations. As the architecture relies on unadulterated pixel shifts to compute optical flow, it is highly susceptible to environmental movement. In the absence of a mathematical camera stabilisation module, the system remains vulnerable to extreme wind shear, which could induce camera oscillations that simulate vehicular kinetic activity. Furthermore, the implementation of a rigid noise floor may inadvertently suppress the kinetic signatures associated with very low-speed, minor collisions (e.g., bumper impacts in parking areas), thereby restricting operational effectiveness to higher-velocity carriageway scenarios. Additionally, whilst the system maintained a throughput of 60 frames per second, the evaluation was performed on pre-recorded, offline video files. The transmission and processing of high-definition RTSP streams from live IP cameras over municipal networks may introduce uncontrollable network latency, which was not considered during this benchmarking process.
\section{Recommendations for Future Work}
To transform this prototype into a fully autonomous deployment at a municipal scale, multiple avenues for future research are recommended. Primarily, the static noise floor should be supplanted by a dynamic, deep-learning-based camera stabilisation system. Incorporating a lightweight convolutional module capable of automatically mapping and compensating for global frame shifts prior to optical flow extraction would substantially diminish false positives, particularly in environments subject to high winds.

Furthermore, the integration of multi-modal sensor fusion would significantly improve the system's performance under zero-visibility conditions, where even dense optical flow methods are ineffective. Finally, as discussed in the literature review, the progression towards incorporating Vision-Language Models (VLMs) into the terminal alert logic presents a highly promising avenue. Rather than generating a simple binary "Accident Detected" indicator, a VLM could semantically analyse the kinetic data streams to produce a descriptive, natural language narrative for emergency responders, thereby substantially reducing the cognitive burden on human operators and expediting the management of critical incidents.

\section{Final Remarks}
The imperative for automated and highly dependable traffic monitoring is set to become increasingly prominent as urban populations continue to grow. This research demonstrates that the implementation of solely spatiotemporal kinetic mapping provides a highly robust, privacy-conscious, and computationally feasible solution. By bridging the gap between theoretical deep learning models and practical edge computing deployment, this framework represents a significant advancement towards the realisation of safer, more intelligent, and more responsive transportation networks.
%=== END OF CHAPTER FIVE ===
\end{spacing}
\newpage
