%=== FRONT PART ===
%=== ABSTRACT ===
\newpage

\chapter*{\centering Abstract}
\markboth{Abstract}{}
\vspace{-0.2in}

\begin{spacing}{1.5}
\setlength{\parskip}{0.3in}
\setlength{\parindent}{0in}

\addcontentsline{toc}{chapter}{Abstract}

The rapid expansion of urban infrastructure necessitates the deployment of robust Intelligent Transport Systems (ITS) to mitigate traffic accidents and congestion. However, conventional vision-based monitoring systems often encounter difficulties in achieving precise spatiotemporal detection of complex collisions whilst satisfying the stringent latency requirements inherent to real-time edge computing; these systems frequently underperform due to their reliance on fragile spatial object tracking. To address this challenge, this study introduces a novel, purely spatiotemporal vision-based framework that entirely omits bounding-box vehicle tracking. The asynchronous architecture assesses unstructured physical motion by extracting Farnebäck dense optical flow, subsequently mapping this raw kinetic data into HSV tensors for probabilistic classification via a custom-trained ResNet-18 Convolutional Neural Network.

The system's performance, assessed against a mathematically balanced synthesis of the CADP and TAD benchmark datasets, exhibited an impressive overall accuracy of 96.61\% and a perfectly balanced F1-score of 0.966. The utilisation of a Binary Cross-Entropy (BCE) loss function effectively distinguished anomalous structural impacts from routine traffic flow without exhibiting a default bias towards the majority class. Significantly, the dual-threaded producer-consumer architecture maintained a real-time processing rate of 60 frames per second on standard consumer edge hardware, achieving a near-instantaneous Time-to-Accident (TTA) verification latency of merely 220 milliseconds. By quantifying pure kinetic energy instead of discrete spatial identifiers, the framework offers a highly resilient, privacy-conscious prototype optimally suited for deployment within contemporary municipal traffic surveillance networks.

\par
% \textbf{Keywords:} Dissertation, keywords.
\end{spacing}
\newpage
%=== END OF ABSTRACT ===
